---
title: "Analyzing HINTS 6 data using R"
author: "Jacob Rohde"
date: '2024-09-05'
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading required libraries and data into R

```{r load packages, message = FALSE, warning = FALSE, results = 'hide'}
library(haven) # For loading data from SAS, SPSS, or STATA into R
library(dplyr) # For data manipulation
library(survey) # For analyzing complex survey data
library(srvyr) # For manipulating survey objects with dplyr
library(broom) # For presenting tidy data tables
library(rstudioapi) # For setting a working directory

# Setting the working directory to file location
setwd(dirname(getActiveDocumentContext()$path))

# Load data
df = haven::read_sas("hints6_public.sas7bdat") 

```

# Recoding survey variables

The code chunk below shows an example of how to use 'dplyr' to create new variables or recode existing ones.

```{r re-coding data}
df = df |>
  dplyr::mutate(gender = case_match(factor(BirthGender),
                                    '1' ~ 'Male',
                                    '2' ~ 'Female')) |>
  
  dplyr::mutate(edu = case_match(factor(Education),
                                 c('1', '2') ~ 'Less than high school',
                                 '3' ~ '12 years or completed high school',
                                 c('4', '5') ~ 'Some college',
                                 c('6', '7') ~ 'College graduate or higher')) |>
  
  dplyr::mutate(SeekCancerInfo = case_match(SeekCancerInfo,
                                             1 ~ 1,
                                             2 ~ 0))

# Setting the reference level for categorical variables
df$gender = relevel(factor(df$gender, ordered = F), 
                    ref = 'Male')

df$edu = relevel(factor(df$edu, ordered = F), 
                 ref = 'Less than high school')

```

# Example analytic procedures using a replicate weights approach

The code chunk below creates a survey design object to account for replicate weights when running statistical analyses.

```{r replicate weights design object}
svy_obj_rep = as_survey_rep(.data = df,
                            weights = PERSON_FINWT0,
                            repweights = num_range(prefix = "PERSON_FINWT", 
                                                   range = 1:50),
                            type = "JKn", 
                            scale = 0.98, 
                            rscales = rep(1, times = 50))
```

**Computing a crosstab and chi-square test:**

```{r replicate weights crosstab and chi-square}
# Crosstab
svy_obj_rep |>
  dplyr::filter(is.na(edu) == F, 
                is.na(gender) == F) |>
  dplyr::group_by(edu, gender) |>
  dplyr::summarize(n = n(),
                   total = survey_total(),
                   pct = survey_prop()) 

# Chi-square test
svy_obj_rep |>
  svychisq(formula = ~ gender + edu,
           statistic = "F")
```

**Computing a logistic regression:**

```{r replicate weights logistic regression}
logistic_model = svy_obj_rep |>
  svyglm(formula = SeekCancerInfo ~ edu + gender,
         family = quasibinomial()) 

# For displaying general summary statistics 
summary(logistic_model) 

# For displaying odds ratios and 95% confidence intervals
tidy(logistic_model, 
     conf.int = T, 
     conf.level = 0.95, 
     exponentiate = T) 
```

**Computing a linear regression:**

```{r replicate weights linear regression}
linear_model = svy_obj_rep |>
  svyglm(formula = GeneralHealth ~ edu + gender,
         family = gaussian()) 

summary(linear_model)
```

# Example analytic procedures using a Taylor Series linearization approach

The code chunk below creates a survey design object to account for Taylor Series linearization sample weights when running statistical analyses.

```{r TS design object}
svy_obj_linear = as_survey_design(.data = df,
                                  ids = VAR_CLUSTER,
                                  strata = VAR_STRATUM,
                                  weights = PERSON_FINWT0,
                                  nest = T)
```

**Computing a crosstab and chi-square test:**

```{r TS crosstab and chi-square}
# Crosstab
svy_obj_linear |>
  dplyr::filter(is.na(edu) == F, 
                is.na(gender) == F) |>
  dplyr::group_by(edu, gender) |>
  dplyr::summarize(n = n(),
                   total = survey_total(),
                   pct = survey_prop()) 

# Chi-square test
svy_obj_linear |>
  svychisq(formula = ~ gender + edu,
           statistic = "F")
```

**Computing a logistic regression:**

```{r TS logistic regression}
logistic_model = svy_obj_linear |>
  svyglm(formula = SeekCancerInfo ~ edu + gender,
         family = quasibinomial()) 

# For displaying general summary statistics 
summary(logistic_model)  

# For displaying odds ratios and 95% confidence intervals
tidy(logistic_model, 
     conf.int = T, 
     conf.level = 0.95, 
     exponentiate = T) 
```

**Computing a linear regression:**

```{r TS linear regression}
linear_model = svy_obj_linear |>
  svyglm(formula = GeneralHealth ~ edu + gender,
         family = gaussian()) 

summary(linear_model)
```

# Combining HINTS 5 Cycle 4 data with HINTS 6

The code chunk below loads HINTS 6 and HINTS 5 Cycle 4 SAS files into R as separate data objects (make sure both files are in the same working directory).

```{r loading H6 and H5C4 data}
# HINTS 6 file
df_H6 = haven::read_sas("hints6_public.sas7bdat")

# HINTS 5 Cycle 4 file
df_H5C4 = haven::read_sas("hints5_cycle4_public.sas7bdat")
```

**Create new sample weights and merge the two data sets:**

```{r merging H6 and H5C4}

# Create variable names
nwgt_var_names = c(paste0('nwgt', 1:100))
var_names = c(paste0('PERSON_FINWT', 1:50))

# Create Hints 5 Cycle 4 group weights
df_H5C4 = df_H5C4 |>
  dplyr::mutate(hints_edition = 'Hints 5 Cycle 4') |>
  dplyr::mutate(nwgt0 = PERSON_FINWT0)

for(i in 1:100){
  if(i <= 50){
    df_H5C4[nwgt_var_names[i]] = df_H5C4[var_names[i]]
  }
  
  if(i > 50){
    df_H5C4[nwgt_var_names[i]] = df_H5C4$PERSON_FINWT0
  }
}

# Create Hints 6 group weights
df_H6 = df_H6 |>
  dplyr::mutate(hints_edition = 'HINTS 6') |> 
  dplyr::mutate(nwgt0 = PERSON_FINWT0)

for(i in 1:100){
  if(i <= 50){
    df_H6[nwgt_var_names[i]] = df_H6$PERSON_FINWT0
  }
  
  if(i > 50){
    df_H6[nwgt_var_names[i]] = df_H6[var_names[i-50]]
  }
}

# Merge the data sets
df_multi = plyr::rbind.fill(df_H5C4, df_H6)

# Display number of respondents from both survey editions
table(df_multi$hints_edition)
```

The example code below can be used to run simple frequencies on two common variables ("SeekCancerInfo" and "ChanceAskQuestions") in the HINTS 6 and HINTS 5 Cycle 4 merged data set using a replicate weights approach:

```{r merged replicate weights design object, results = 'hide'}
# Create the replicate weights survey design object
svy_obj_rep_merged = as_survey_rep(.data = df_multi,
                                   weights = nwgt0,
                                   repweights = num_range(prefix = "nwgt", 
                                                          range = 1:100),
                                   type = "JKn", 
                                   scale = 0.98, 
                                   rscales = rep(1, times = 100))

# Crosstab 
svy_obj_rep_merged |>
  dplyr::filter(ChanceAskQuestions > 0, 
                SeekCancerInfo > 0) |>
  dplyr::group_by(ChanceAskQuestions, SeekCancerInfo) |>
  dplyr::summarize(n = n(),
                   total = survey_total(),
                   pct = survey_prop()) 
```

The example code below can be used to run simple frequencies on two common variables ("SeekCancerInfo" and "ChanceAskQuestions") in the HINTS 6 and HINTS 5 Cycle 4 merged data set using a Taylor Series linearization approach:

```{r merged TS weights design object, results = 'hide'}
# Create the Taylor Series linearization survey design object
svy_obj_linear_merged = as_survey_design(.data = df_multi,
                                         ids = VAR_CLUSTER,
                                         strata = VAR_STRATUM,
                                         weights = PERSON_FINWT0,
                                         nest = T)

# Crosstab 
svy_obj_linear_merged |>
  dplyr::filter(ChanceAskQuestions > 0, 
                SeekCancerInfo > 0) |>
  dplyr::group_by(ChanceAskQuestions, SeekCancerInfo) |>
  dplyr::summarize(n = n(),
                   total = survey_total(),
                   pct = survey_prop()) 
```
